---
title: "06 - Distributions & ggplot2"
author: "Joachim Goedhart"
date: '2022-09-08'
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 4
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 6.1 Setup and data

Let's load the `{tidyverse}` package, which includes the `{ggplot2}`
package that we will use for plotting:

```{r}
  library(tidyverse)
```

For the first set of visualizations, we use the ChickWeight data set
that comes with R. Let's look at the first few lines:

```{r chickweight}
head(ChickWeight)
```

It can be useful to look at the last lines as well:

```{r}
tail(ChickWeight)
```

To summarize the dataset:

```{r}
summary(ChickWeight) #dataset is present in the workspace with the dataset name
```

### 6.2 Multiple distributions

#### 6.3 Plotting the data

You see that you have 4 columns, most importantly weight. There are 3
columns with 'metadata': the Time that a chick weighed, which Chick, and
its Diet. Let's plot the weight versus Time:

```{r chickweight-scatterAll}
ggplot(data = ChickWeight, aes(x=Time, y=weight)) +
  geom_point()
```

Let's change the defauklt theme to set the background of the plot panel
to white:

```{r}
theme_set(theme_light(base_size = 14))
```

The points are overlapping. To reduce overlap, we can introduce random
displacement, also known as jitter. The `width` can be used to control
the amount of jitter:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight)) +
  geom_jitter()
```

Since there is still overlap, the dots can be made more transparent:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight)) +
  geom_jitter(width = 0.25, alpha=0.5)
```

### 6.4 Boxplots

There are several ways to summarize the distribution of the data. The
most common are barplots, boxplots and violinplots. The barplot is an
over simplication of the data that only depicts the average. This type
of summary should be avoided, see
[here](https://doi.org/10.1371/journal.pbio.1002128) for more about
that. Boxplots and violinplots are better suited as summaries.

In a boxplot, the values that are depicted by default are:

-   the median (big bar in the middle = the 50th percentile)
-   the 25th and 75 percentiles (ends of the box = quartiles)
-   the whiskers extend to the last points within 1.5 \* IQR from the
    quartiles (IQR is the interquartile range, i.e. the range between
    the box's ends)
-   all points that lie outside are plotted as points (outliers)

Here we go:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight)) + geom_boxplot()
```

In the plot above, all data is used for the summary, so we should
indicate that we want to group the data by time:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) + geom_boxplot()
```

Since different data distribution can give rise to the same boxplot, it
is still useful to plot the data:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_jitter(width = 0.25, alpha=0.5) +
  geom_boxplot()
```

The boxplot is now on top of the data. This can be solved by changing
the order of the the layers or by changing the fill color of the
boxplot:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_jitter(width = 0.25, alpha=0.5) +
  geom_boxplot(fill=NA)
```

One other thing needs to be fixed. The outliers are plotted twice. Once
as a jittered dot and once as outlier of the boxplot. To remove the
outlier from the boxplot we use:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_jitter(width = 0.25, alpha=0.5) +
  geom_boxplot(fill=NA, outlier.shape = NA)
```

This graph looks a bit funny, since the Time points are not equally
spaced. There is a difference of 2 for almost all, execpt for the last
two time points which are 20 and 21. When there is continuous data on
the x-axis (e.g. time, length, concentration) it makes sense to use
those data for the x-axis position. However, we could also treat the
x-axis data as labels or categories (instead of numbers). This would
result in equal spacing, although the values are not equally spaced. To
do this, we need to convert the Time data to a 'factor'. This can be
done in the dataset or directly in the ggplot command as shown below:

```{r}
ggplot(data = ChickWeight, aes(x=as.factor(Time), y=weight, group=Time)) +
  geom_jitter(width = 0.25, alpha=0.5) +
  geom_boxplot(fill=NA, outlier.shape = NA)
```

Another explanation of discrete versus quantitative data on the x-axis
with ggplot example code [can be found
here](https://joachimgoedhart.github.io/DataViz-protocols/plotting-the-data.html#discrete-conditions).

The boxplot is a summary of the data that shows the midpoint of the data
(median = middle line), the middle 50% of the data (the box) and the
extremes with the whiskers. The exact defenition can vary between the
different whiskers, to it's a good idea to check that. From the
[`geom_boxplot()`
webpage](https://ggplot2.tidyverse.org/reference/geom_boxplot.html):
"The upper whisker extends from the hinge to the largest value no
further than 1.5 \* IQR from the hinge (where IQR is the inter-quartile
range, or distance between the first and third quartiles). The lower
whisker extends from the hinge to the smallest value at most 1.5 \* IQR
of the hinge. Data beyond the end of the whiskers are called "outlying"
points and are plotted individually."

### 6.5 Violinplots

To better summarize the distribution, a violinplot can be displayed:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) + geom_violin(scale = "width", fill="grey")
```

If desired the quantiles can be displayed. These can be user-defined and
here we use the quartiles that corresponds to the horizontal lines that
define the boxplot. So the middle line is the median:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_violin(scale = "width", fill="grey", draw_quantiles = c(0.25,0.5,0.75))
```

Another option is to combine a boxplot with a violinplot to display the
quartiles:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_violin(scale = "width", fill="grey") +
  geom_boxplot(fill="white", outlier.shape = NA, width=0.2)
```

To combine the violinplot with the data:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_violin(scale = "width", fill="grey") +
  geom_jitter(width = 0.25, alpha=0.5)
```

The data are acquired for different Diets so we can map that data to
different colors:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_violin(scale = "width", fill="grey") +
  geom_jitter(width = 0.25, alpha=0.5, aes(color=Diet))
```

### 6.6 Facets: Splitting into multiple plots

Let's say we want to compare the weight for different diets. In ggplot
there is a way to 'split' the data and show it in multiple graphs. The
simplest function is `facet_wrap()` and let's see how that works:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_violin(scale = "width", fill="grey") +
  geom_jitter(width = 0.25, alpha=0.5, aes(color=Diet)) +
  facet_wrap(~Diet)
```

The `facet_wrap()` function calculates the optimal layout, but it can be
defined by ncol (or nrow):

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_violin(scale = "width", fill="grey") +
  geom_jitter(width = 0.25, alpha=0.5, aes(color=Diet)) +
  facet_wrap(~Diet, ncol = 4)
```

The use of `facet_wrap()` is particularly powerful when the data has
several conditions. Another, similar function, is `facet_grid()` which
allows splitting the data according to two parameters:

```{r}
ggplot(data = ChickWeight, aes(x=Time, y=weight, group=Time)) +
  geom_violin(scale = "width", fill="grey") +
  geom_jitter(width = 0.25, alpha=0.5, aes(color=Diet)) +
  facet_grid(Diet~Time, scales = "free_x") +
  theme(axis.text.x = element_blank()) #Removes the Time labels from the x-axis
```

In the example here, it may not be the best idea to split the Time in
different plots, but it illustrates how `facet_grid()` can be used.

### 6.7 A single distribution

Let's have a look at the combined data of the first and last week from
diet 3. This can be achieved in multiple ways. For instance we can
generate a new dataframe. But we can also filter the data and feed it
directly into ggplot with the pipe operator:

```{r}
ChickWeight %>% filter(Diet==3) %>%
  filter(Time %in% c(0,2,4,6,16,18,20,21)) %>%
  ggplot(aes(x=Diet, y=weight)) +
  geom_violin(scale = "width", fill="grey90") +
  geom_jitter(width = 0.25, alpha=1, size=5)
```

To generate a new dataframe:

```{r}
ChickWeight_sub <- ChickWeight %>%
  filter(Diet==3) %>%
  filter(Time %in% c(0,2,4,6,16,18,20,21))
```

#### 6.7.1 A single violinplot

Since we have combined data from the first and last week, we see a bimodal
distribution. This is nicely visualized with the violinplot. The
violinplot shows the distribution of the data between the two extremes.
The violinplot shows the *estimated* density distribution of the data.
This ranges from the minimal to the maximal value of the data. To
extrapolate the distribution we can use `trim=FALSE`:

```{r}
ggplot(ChickWeight_sub, aes(x=Diet, y=weight)) +
  geom_violin(scale = "width", fill="grey90", trim = FALSE) +
  geom_jitter(width = 0.25, alpha=1, size=5)
```

#### 6.7.2 Density plot

The violinplot is much like a smoothened hisogram. We can plot one side
of the violinplot and in this case it is called a density plot:

```{r}
ggplot(ChickWeight_sub, aes(y=weight)) +
  geom_density(fill="grey90")
```

These plots are typically rotated 90 degrees:

```{r}
ggplot(ChickWeight_sub, aes(x=weight)) +
  geom_density(fill="grey90")
```

To add the data we can use `geom_rug()`:

```{r}
ggplot(ChickWeight_sub, aes(x=weight)) +
  geom_density(fill="grey90") +
  geom_rug()
```

#### 6.7.3 Histogram

To display these data as a histogram instead:

```{r}
ggplot(ChickWeight_sub, aes(x=weight)) +
  geom_histogram()
```

And we can combine them:

```{r}
ggplot(ChickWeight_sub, aes(x=weight)) +
  geom_histogram() + geom_density()
```

The density estimate is invisible, since it has a different amplitude.
To solve this:

```{r}
# From: https://stackoverflow.com/questions/37404002/geom-density-to-match-geom-histogram-binwitdh
ggplot(ChickWeight_sub, aes(x=weight)) +
  geom_histogram(binwidth = 10, fill="grey80") +
  geom_density(aes(y = 10*..count..), fill="black", alpha=0.5)
```

One thing that is not so nice is the margin between the plot and the
axis. To remove the margin:

```{r}
# From: https://stackoverflow.com/questions/37404002/geom-density-to-match-geom-histogram-binwitdh
ggplot(ChickWeight_sub, aes(x=weight)) +
  geom_histogram(binwidth = 10, fill="grey80") +
  geom_density(aes(y = 10*..count..), fill="black", alpha=0.5) +
  scale_y_continuous(expand = c(0,NA))
```

To further improve this, we can set the maximal value of the x-axis:

```{r}
# From: https://stackoverflow.com/questions/37404002/geom-density-to-match-geom-histogram-binwitdh
ggplot(ChickWeight_sub, aes(x=weight)) +
  geom_histogram(binwidth = 10, fill="grey80") +
  geom_density(aes(y = 10*..count..), fill="black", alpha=0.5) +
  scale_y_continuous(expand = c(0,NA), limits = c(0,11))
```

### 6.8 Two distributions

#### 6.8.1 Histogram & Density

It is possible to overlay multiple histograms and/or density curves:

```{r}
ChickWeight %>% filter(Diet %in% c(2,3)) %>%
  filter(Time %in% c(0,2,4,6,16,18,20,21)) %>%
  ggplot(aes(x=weight, fill=Diet)) +
  geom_histogram(binwidth = 10) +
  geom_density(aes(y = 10*..count..), alpha=0.5) +
  scale_y_continuous(expand = c(0,NA), limits = c(0,21))
```

By default the histograms are stacked (unlike the density plot), making
it difficult to compare them. This can be solved and requires
tranparancy because they partially overlap.

```{r}
ChickWeight %>% filter(Diet %in% c(2,3)) %>%
  filter(Time %in% c(0,2,4,6,16,18,20,21)) %>%
  ggplot(aes(x=weight, fill=Diet)) +
  geom_histogram(binwidth = 10, alpha=0.8, position="identity") +
  geom_density(aes(y = 10*..count..), alpha=0.5) +
  scale_y_continuous(expand = c(0,NA), limits = c(0,11))
```

#### 6.8.2 Splitting into two plots

However, overlaying histograms is usually not a great idea as this
generates a cluttered plot. One way to solve this is by splitting the
data with `facet_wrap()`. We can use `labeller` to use both the
parameter as its value for labeling the plots. In this case the legend
can be removed as it is redundant, allowing more space for the data:

```{r}
ChickWeight %>% filter(Diet %in% c(2,3)) %>%
  filter(Time %in% c(0,2,4,6,16,18,20,21)) %>%
  ggplot(aes(x=weight, fill=Diet)) +
  geom_histogram(binwidth = 10, alpha=0.8, position="identity") +
  geom_density(aes(y = 10*..count..), alpha=0.5) +
  scale_y_continuous(expand = c(0,NA), limits = c(0,11)) +
  facet_wrap(~Diet, labeller = label_both) +
  theme(legend.position = "none")
```
