---
title: "07 - Heatmaps *et al.* - base R"
author: "Anna Heintz-Buschart"
date: '2022-09-07'
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 4
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 7.1 Setup

In this notebook, we take a look at R for visualizing data matrices.

```{r libraries}
if(!require(RColorBrewer)){
  install.packages("RColorBrewer",repos = "http://cran.us.r-project.org")
  library(RColorBrewer)
}
if(!require(gplots)){
  install.packages("gplots",repos = "http://cran.us.r-project.org")
  library(gplots)
}
if(!require(gtools)){
  install.packages("gtools",repos = "http://cran.us.r-project.org")
  library(gtools)
}
```

Heatmaps are a commonly used visualization tool for large data matrices,
where you'd have many features (e.g. gene's expression values, microbial
abundances, metabolite levels...) measured over a bunch of samples.

### 7.2 Loading data

In this example, we'll work with a metabolomics dataset from a [study on
gastectomy](https://gut.bmj.com/content/69/8/1404), processed for a
[meta-analysis](https://doi.org/10.1186/s40168-021-01149-z). There is
also some metadata available.

```{r dataset}
metabo <- read.delim("ERAWIJANTARI_metab.tsv",row.names=1) #data
colnames(metabo) <- sub("^X","",colnames(metabo))
metabo <- as.matrix(metabo)
metabo_log <- log10(metabo)
metaD <- read.delim("ERAWIJANTARI_metaD.tsv") #metadata
all(colnames(metabo) == metaD$Sample) #check that samples match in data matrices
```

With 96 samples and 211 metabolites, this is not a huge data set. But
it's awkward to analyse and visualize bit by bit. A heatmap can give a
quick overview.

### 7.3 Basic heatmaps
#### 7.3.1 A simple heatmap

Here's how the default heatmap looks:

```{r heatmapBasic}
heatmap(metabo)
```

Every column represents one sample, every row represents a metabolite.
Let's for a moment ignore that there are too many datapoints to plot all
the labels - indeed, let's ignore the labels for now:

```{r heatmapBasic-text}
heatmap(metabo,
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5)) # remove the white space around
```

#### 7.3.2 Scaling

What this simple visualization does not tell you is what the colours
mean - we'll come back to that later - but the darker colours are higher
values. You see that there are some high values for all metabolites.
When you think about it, this must mean that something was done to our
data, because you would never find all metabolites at a similar level.
This is true, the default setting for a heatmap is to scale **by row**.
This means that the average value per row has the middle colour, and the
further a value is away from the average, the lighter or darker it will
be. How light or dark? scaling takes the distribution into account, so
all values that are one standard deviation higher than its row's average
have the same colour. You can turn scaling off:

```{r heatmapBasic-noScale}
heatmap(metabo,
        scale = "none", # no scaling!
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5)) # remove the white space around
```

Now, you've lost most information, except that you can see that there
are really very few metabolites with a high value. It's not useful not
to scale for this kind of data, but if all your rows have similar
values, it can be better. You can see this, if you use the
log-transformed version of the data set, where all values sit somewhere
between -1 and 6:

You could also scale by column:

```{r heatmapBasic-log-noScale}
heatmap(metabo_log,
        scale = "none", # no scaling!
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5)) # remove the white space around
```

Instant patterns! *As an exercise, plot the heatmap for this data set
with row scaling (`scale = "row"`)*.

When comparing no scaling and row-scaling, you see that no scaling
emphasizes the absolute values, while row-scaling emphasizes differences
between samples. It depends on your question which visualization is
appropriate.

```{r heatmapBasic-log-colScale}
heatmap(metabo_log,
        scale = "col", # column scaling!
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5)) # remove the white space around
```

Now some of the columns which were lighter before cannot be
distinguished anymore. So, you see, scaling is important for the visual
impression.

Generally, we would *recommend* that you transform and normalize (and
potentially scale/centre) your data set appropriately, before you
visualize it. That way, your analyses are performed correctly and on the
thing you see. What exactly is "appropriate" depends on your data set.
Here, we're just learning visualization, so we'll stick with the
un-normalized, un-scaled/centred (but log-transformed) data.

#### 7.3.3 Colour schemes

A lot of the information in heatmaps comes from the colours. So it's
important to choose good colors. Generally speaking, if the data is not
centred/scaled, use light colours for low values and darker colours for
high values. If you are plotting a scaled/centred data set, it's good
practice to have a neutral color (black or whiteish) in the middle and
stronger colours for the extremes:

```{r heatmapBasic-log-rowScale-colours}
heatmap(metabo_log,
        col = hcl.colors(12,"RdYlBu", rev = TRUE), #diverging palette
        scale = "row", # row scaling!
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5)) # remove the white space around
```

It would be nice to have a colour bar here. So, let's use a bit more
fancy function - which also centers the colours:

```{r heatmap2-log-rowScale-colours}
heatmap.2(metabo_log,
        col = hcl.colors(13,"RdYlBu", rev = TRUE), #diverging palette
        scale = "row", # row scaling!
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5), # remove the white space around
        trace = "none", #don't plot a trace
        density.info = "none") #don't overlay a histogram in the colour bar
```

13 colour steps were not really enough (R's base `heatmap` function uses
only 12...), so here we set it to 255, giving a smooth colour scale:

```{r heatmap2-log-rowScale-MoreColours}
heatmap.2(metabo_log,
        col = hcl.colors(255,"RdYlBu", rev = TRUE), #diverging palette - 255 colours!
        scale = "row", # row scaling
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5), # remove the white space around
        trace = "none", #don't plot a trace
        density.info = "none") #don't overlay a histogram in the colour bar
```

This would be the un-scaled image:

```{r heatmap2-log-noScale-MoreColours}
heatmap.2(metabo_log,
       col = hcl.colors(255,"YlGnBu", rev = TRUE), #single colour palette! - 255 colours
        symbreaks = F, #colours are not centred around 0
        scale = "none", # no scaling!
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5), # remove the white space around
        trace = "none", #don't plot a trace
        density.info = "none") #don't overlay a histogram in the colour bar
```

The colour key is a bit buggy, but Fortunately, it's R, so we can change
the function. Here's the code (not displayed in the .html).

```{r heatmapFunction, include=FALSE}
heatmap.2a <- function (x, Rowv = TRUE, Colv = if (symm) "Rowv" else TRUE, 
                        distfun = dist, hclustfun = hclust, dendrogram = c("both", 
                                                                           "row", "column", "none"), symm = FALSE, 
                        scale = c("none", "row", "column"), na.rm = TRUE, revC = identical(Colv, "Rowv"), add.expr, breaks, 
                        symbreaks = min(x < 0, na.rm = TRUE) || scale != "none", col = "heat.colors", colsep, rowsep, 
                        sepcolor = "white", sepwidth = c(0.05, 0.05), cellnote, notecex = 1, 
                        notecol = "cyan", na.color = par("bg"), trace = c("column", "row", "both", "none"), tracecol = "cyan", hline = median(breaks), 
                        vline = median(breaks), linecol = tracecol, margins = c(5,5), ColSideColors, RowSideColors, cexRow = 0.2 + 1/log10(nr), 
                        cexCol = 0.2 + 1/log10(nc), labRow = NULL, labCol = NULL, srtRow = NULL, srtCol = NULL, adjRow = c(0, NA), adjCol = c(NA,0),
                        offsetRow = 0.5, offsetCol = 0.5, key = TRUE, keysize = 1.5, 
                        density.info = c("histogram", "density", "none"), denscol = tracecol, 
                        symkey = min(x < 0, na.rm = TRUE) || symbreaks, densadj = 0.25, 
                        main = NULL, xlab = NULL, ylab = NULL, lmat = NULL, lhei = NULL, 
                        lwid = NULL, extrafun = NULL, font.col=1, font.row=1, keyName="Value", 
                        ColSideFac=1, RowSideFac=1, ...) 
{
  scale01 <- function(x, low = min(x), high = max(x)) {
    x <- (x - low)/(high - low)
    x
  }
  retval <- list()
  scale <- if (symm && missing(scale)) 
    "none"
  else match.arg(scale)
  dendrogram <- match.arg(dendrogram)
  trace <- match.arg(trace)
  density.info <- match.arg(density.info)
  if (length(col) == 1 && is.character(col)) 
    col <- get(col, mode = "function")
  if (!missing(breaks) && (scale != "none")) 
    warning("Using scale=\"row\" or scale=\"column\" when breaks are", 
            "specified can produce unpredictable results.", "Please consider using only one or the other.")
  if (is.null(Rowv) || is.na(Rowv)) 
    Rowv <- FALSE
  if (is.null(Colv) || is.na(Colv)) 
    Colv <- FALSE
  else if (Colv == "Rowv" && !isTRUE(Rowv)) 
    Colv <- FALSE
  if (length(di <- dim(x)) != 2 || !is.numeric(x)) 
    stop("`x' must be a numeric matrix")
  nr <- di[1]
  nc <- di[2]
  if (nr <= 1 || nc <= 1) 
    stop("`x' must have at least 2 rows and 2 columns")
  if (!is.numeric(margins) || length(margins) != 2) 
    stop("`margins' must be a numeric vector of length 2")
  if (missing(cellnote)) 
    cellnote <- matrix("", ncol = ncol(x), nrow = nrow(x))
  if (!inherits(Rowv, "dendrogram")) {
    if (((!isTRUE(Rowv)) || (is.null(Rowv))) && (dendrogram %in% 
                                                 c("both", "row"))) {
      if (is.logical(Colv) && (Colv)) 
        dendrogram <- "column"
      else dedrogram <- "none"
      warning("Discrepancy: Rowv is FALSE, while dendrogram is `", 
              dendrogram, "'. Omitting row dendogram.")
    }
  }
  if (!inherits(Colv, "dendrogram")) {
    if (((!isTRUE(Colv)) || (is.null(Colv))) && (dendrogram %in% 
                                                 c("both", "column"))) {
      if (is.logical(Rowv) && (Rowv)) 
        dendrogram <- "row"
      else dendrogram <- "none"
      warning("Discrepancy: Colv is FALSE, while dendrogram is `", 
              dendrogram, "'. Omitting column dendogram.")
    }
  }
  if (inherits(Rowv, "dendrogram")) {
    ddr <- Rowv
    rowInd <- order.dendrogram(ddr)
  }
  else if (is.integer(Rowv)) {
    hcr <- hclustfun(distfun(x))
    ddr <- as.dendrogram(hcr)
    ddr <- reorder(ddr, Rowv)
    rowInd <- order.dendrogram(ddr)
    if (nr != length(rowInd)) 
      stop("row dendrogram ordering gave index of wrong length")
  }
  else if (isTRUE(Rowv)) {
    Rowv <- rowMeans(x, na.rm = na.rm)
    hcr <- hclustfun(distfun(x))
    ddr <- as.dendrogram(hcr)
    ddr <- reorder(ddr, Rowv)
    rowInd <- order.dendrogram(ddr)
    if (nr != length(rowInd)) 
      stop("row dendrogram ordering gave index of wrong length")
  }
  else {
    rowInd <- nr:1
  }
  if (inherits(Colv, "dendrogram")) {
    ddc <- Colv
    colInd <- order.dendrogram(ddc)
  }
  else if (identical(Colv, "Rowv")) {
    if (nr != nc) 
      stop("Colv = \"Rowv\" but nrow(x) != ncol(x)")
    if (exists("ddr")) {
      ddc <- ddr
      colInd <- order.dendrogram(ddc)
    }
    else colInd <- rowInd
  }
  else if (is.integer(Colv)) {
    hcc <- hclustfun(distfun(if (symm) 
      x
      else t(x)))
    ddc <- as.dendrogram(hcc)
    ddc <- reorder(ddc, Colv)
    colInd <- order.dendrogram(ddc)
    if (nc != length(colInd)) 
      stop("column dendrogram ordering gave index of wrong length")
  }
  else if (isTRUE(Colv)) {
    Colv <- colMeans(x, na.rm = na.rm)
    hcc <- hclustfun(distfun(if (symm) 
      x
      else t(x)))
    ddc <- as.dendrogram(hcc)
    ddc <- reorder(ddc, Colv)
    colInd <- order.dendrogram(ddc)
    if (nc != length(colInd)) 
      stop("column dendrogram ordering gave index of wrong length")
  }
  else {
    colInd <- 1:nc
  }
  retval$rowInd <- rowInd
  retval$colInd <- colInd
  retval$call <- match.call()
  x <- x[rowInd, colInd]
  x.unscaled <- x
  cellnote <- cellnote[rowInd, colInd]
  if (is.null(labRow)) 
    labRow <- if (is.null(rownames(x))) 
      (1:nr)[rowInd]
  else rownames(x)
  else labRow <- labRow[rowInd]
  if (is.null(labCol)) 
    labCol <- if (is.null(colnames(x))) 
      (1:nc)[colInd]
  else colnames(x)
  else labCol <- labCol[colInd]
  if (scale == "row") {
    retval$rowMeans <- rm <- rowMeans(x, na.rm = na.rm)
    x <- sweep(x, 1, rm)
    retval$rowSDs <- sx <- apply(x, 1, sd, na.rm = na.rm)
    x <- sweep(x, 1, sx, "/")
  }
  else if (scale == "column") {
    retval$colMeans <- rm <- colMeans(x, na.rm = na.rm)
    x <- sweep(x, 2, rm)
    retval$colSDs <- sx <- apply(x, 2, sd, na.rm = na.rm)
    x <- sweep(x, 2, sx, "/")
  }
  if (missing(breaks) || is.null(breaks) || length(breaks) < 
      1) {
    if (missing(col) || is.function(col)) 
      breaks <- 16
    else breaks <- length(col) + 1
  }
  if (length(breaks) == 1) {
    if (!symbreaks) 
      breaks <- seq(min(x, na.rm = na.rm), max(x, na.rm = na.rm), 
                    length = breaks)
    else {
      extreme <- max(abs(x), na.rm = TRUE)
      breaks <- seq(-extreme, extreme, length = breaks)
    }
  }
  nbr <- length(breaks)
  ncol <- length(breaks) - 1
  if (class(col) == "function") 
    col <- col(ncol)
  min.breaks <- min(breaks)
  max.breaks <- max(breaks)
  x[x < min.breaks] <- min.breaks
  x[x > max.breaks] <- max.breaks
  if (missing(lhei) || is.null(lhei)) 
    lhei <- c(keysize, 4)
  if (missing(lwid) || is.null(lwid)) 
    lwid <- c(keysize, 4)
  if (missing(lmat) || is.null(lmat)) {
    lmat <- rbind(4:3, 2:1)
    if (!missing(ColSideColors)) {
      if (is.vector(ColSideColors)) ColSideColors <- as.matrix(t(ColSideColors))
      if (!is.character(ColSideColors) || ncol(ColSideColors) != nc) 
        stop("'ColSideColors' must be a character vector of length ncol(x) or character matrix with ncol ncol(x)")
      lmat <- rbind(lmat[1, ] + 1, c(NA, 1), lmat[2, ] + 1)
      lhei <- c(lhei[1], 0.12*nrow(ColSideColors)*ColSideFac, lhei[2])
    }
    if (!missing(RowSideColors)) {
      if (is.vector(RowSideColors)) RowSideColors <- as.matrix(RowSideColors)
      if (!is.character(RowSideColors) || nrow(RowSideColors) != nr) 
        stop("'RowSideColors' must be a character vector of length nrow(x) or character matrix with nrow nrow(x)")
      lmat <- cbind(lmat[, 1] + 1, c(rep(NA, nrow(lmat) -  1), 1), lmat[, 2] + 1)
      lwid <- c(lwid[1], 0.12*ncol(RowSideColors)*RowSideFac, lwid[2])
    }
    lmat[is.na(lmat)] <- 0
  }
  if (length(lhei) != nrow(lmat)) 
    stop("lhei must have length = nrow(lmat) = ", nrow(lmat))
  if (length(lwid) != ncol(lmat)) 
    stop("lwid must have length = ncol(lmat) =", ncol(lmat))
  op <- par(no.readonly = TRUE)
  on.exit(par(op))
  layout(lmat, widths = lwid, heights = lhei, respect = FALSE)
  if (!missing(RowSideColors)) {
    par(mar = c(margins[1], 0, 0, 0.5))
    image(matrix(1:(ncol(RowSideColors)*nr),ncol=nr,nrow=ncol(RowSideColors),byrow=F), col = t(RowSideColors[rowInd,]), axes = FALSE)
  }
  if (!missing(ColSideColors)) {
    par(mar = c(0.5, 0, 0, margins[2]))
    image(matrix(1:(nrow(ColSideColors)*nc),nrow=nc,ncol=nrow(ColSideColors),byrow=F),
          col=t(ColSideColors[,colInd]),axes = FALSE)
  }
  par(mar = c(margins[1], 0, 0, margins[2]))
  x <- t(x)
  cellnote <- t(cellnote)
  if (revC) {
    iy <- nr:1
    if (exists("ddr")) 
      ddr <- rev(ddr)
    x <- x[, iy]
    cellnote <- cellnote[, iy]
  }
  else iy <- 1:nr
  image(1:nc, 1:nr, x, xlim = 0.5 + c(0, nc), ylim = 0.5 + 
          c(0, nr), axes = FALSE, xlab = "", ylab = "", col = col, 
        breaks = breaks, ...)
  retval$carpet <- x
  if (exists("ddr")) 
    retval$rowDendrogram <- ddr
  if (exists("ddc")) 
    retval$colDendrogram <- ddc
  retval$breaks <- breaks
  retval$col <- col
  if (!invalid(na.color) & any(is.na(x))) {
    mmat <- ifelse(is.na(x), 1, NA)
    image(1:nc, 1:nr, mmat, axes = FALSE, xlab = "", ylab = "", 
          col = na.color, add = TRUE)
  }
  if (is.null(srtCol)) 
    axis(1, 1:nc, labels = labCol, las = 2, line = -0.5 + 
           offsetCol, tick = 0, cex.axis = cexCol, hadj = adjCol[1], 
         padj = adjCol[2], font=font.col)
  else {
    if (is.numeric(srtCol)) {
      if (missing(adjCol) || is.null(adjCol)) 
        adjCol = c(1, NA)
      xpd.orig <- par("xpd")
      par(xpd = NA)
      xpos <- axis(1, 1:nc, labels = rep("", nc), las = 2, 
                   tick = 0)
      text(x = xpos, y = par("usr")[3] - (1 + offsetCol) * 
             strheight("M"), labels = labCol, adj = adjCol, 
           cex = cexCol, srt = srtCol, font=font.col)
      par(xpd = xpd.orig)
    }
    else warning("Invalid value for srtCol ignored.")
  }
  if (is.null(srtRow)) {
    axis(4, iy, labels = labRow, las = 2, line = -0.5 + offsetRow, 
         tick = 0, cex.axis = cexRow, hadj = adjRow[1], padj = adjRow[2], font=font.row)
  }
  else {
    if (is.numeric(srtRow)) {
      xpd.orig <- par("xpd")
      par(xpd = NA)
      ypos <- axis(4, iy, labels = rep("", nr), las = 2, 
                   line = -0.5, tick = 0)
      text(x = par("usr")[2] + (1 + offsetRow) * strwidth("M"), 
           y = ypos, labels = labRow, adj = adjRow, cex = cexRow, 
           srt = srtRow, font=font.row)
      par(xpd = xpd.orig)
    }
    else warning("Invalid value for srtRow ignored.")
  }
  if (!is.null(xlab)) 
    mtext(xlab, side = 1, line = margins[1] - 1.25)
  if (!is.null(ylab)) 
    mtext(ylab, side = 4, line = margins[2] - 1.25)
  if (!missing(add.expr)) 
    eval(substitute(add.expr))
  if (!missing(colsep)) 
    for (csep in colsep) rect(xleft = csep + 0.5, ybottom = 0, 
                              xright = csep + 0.5 + sepwidth[1], ytop = ncol(x) + 
                                1, lty = 1, lwd = 1, col = sepcolor, border = sepcolor)
  if (!missing(rowsep)) 
    for (rsep in rowsep) rect(xleft = 0, ybottom = (ncol(x) + 
                                                      1 - rsep) - 0.5, xright = nrow(x) + 1, ytop = (ncol(x) + 
                                                                                                       1 - rsep) - 0.5 - sepwidth[2], lty = 1, lwd = 1, 
                              col = sepcolor, border = sepcolor)
  min.scale <- min(breaks)
  max.scale <- max(breaks)
  x.scaled <- scale01(t(x), min.scale, max.scale)
  if (trace %in% c("both", "column")) {
    retval$vline <- vline
    vline.vals <- scale01(vline, min.scale, max.scale)
    for (i in colInd) {
      if (!is.null(vline)) {
        abline(v = i - 0.5 + vline.vals, col = linecol, 
               lty = 2)
      }
      xv <- rep(i, nrow(x.scaled)) + x.scaled[, i] - 0.5
      xv <- c(xv[1], xv)
      yv <- 1:length(xv) - 0.5
      lines(x = xv, y = yv, lwd = 1, col = tracecol, type = "s")
    }
  }
  if (trace %in% c("both", "row")) {
    retval$hline <- hline
    hline.vals <- scale01(hline, min.scale, max.scale)
    for (i in rowInd) {
      if (!is.null(hline)) {
        abline(h = i - 0.5 + hline.vals, col = linecol, 
               lty = 2)
      }
      yv <- rep(i, ncol(x.scaled)) + x.scaled[i, ] - 0.5
      yv <- rev(c(yv[1], yv))
      xv <- length(yv):1 - 0.5
      lines(x = xv, y = yv, lwd = 1, col = tracecol, type = "s")
    }
  }
  if (!missing(cellnote)) 
    text(x = c(row(cellnote)), y = c(col(cellnote)), labels = c(cellnote), 
         col = notecol, cex = notecex)
  par(mar = c(margins[1], 0, 0, 0))
  if (dendrogram %in% c("both", "row")) {
    plot(ddr, horiz = TRUE, axes = FALSE, yaxs = "i", leaflab = "none")
  }
  else plot.new()
  par(mar = c(0, 0, if (!is.null(main)) 5 else 0, margins[2]))
  if (dendrogram %in% c("both", "column")) {
    plot(ddc, axes = FALSE, xaxs = "i", leaflab = "none")
  }
  else plot.new()
  if (!is.null(main)) 
    title(main, cex.main = 1.5 * op[["cex.main"]])
  if (key) {
    par(mar = c(5, 4, 2, 1), cex = 0.75)
    tmpbreaks <- breaks
    if (symkey) {
      max.raw <- max(abs(c(x, breaks)), na.rm = TRUE)
      min.raw <- -max.raw
      tmpbreaks[1] <- -max(abs(x), na.rm = TRUE)
      tmpbreaks[length(tmpbreaks)] <- max(abs(x), na.rm = TRUE)
    }
    else {
      min.raw <- min(x, na.rm = TRUE)
      max.raw <- max(x, na.rm = TRUE)
    }
    z <- seq(min.raw, max.raw, length = length(col))
    image(z = matrix(z, ncol = 1), col = col, breaks = tmpbreaks, 
          xaxt = "n", yaxt = "n")
    par(usr = c(0, 1, 0, 1))
    lv <- pretty(breaks)
    xv <- scale01(as.numeric(lv), min.raw, max.raw)
    axis(1, at = xv, labels = lv)
    if (scale == "row") 
      mtext(side = 1, "Row Z-Score", line = 2)
    else if (scale == "column") 
      mtext(side = 1, "Column Z-Score", line = 2)
    else mtext(side = 1, keyName, line = 2)
    if (density.info == "density") {
      dens <- density(x, adjust = densadj, na.rm = TRUE)
      omit <- dens$x < min(breaks) | dens$x > max(breaks)
      dens$x <- dens$x[-omit]
      dens$y <- dens$y[-omit]
      dens$x <- scale01(dens$x, min.raw, max.raw)
      lines(dens$x, dens$y/max(dens$y) * 0.95, col = denscol, 
            lwd = 1)
      axis(2, at = pretty(dens$y)/max(dens$y) * 0.95, pretty(dens$y))
      title("Color Key\nand Density Plot")
      par(cex = 0.5)
      mtext(side = 2, "Density", line = 2)
    }
    else if (density.info == "histogram") {
      h <- hist(x, plot = FALSE, breaks = breaks)
      hx <- scale01(breaks, min.raw, max.raw)
      hy <- c(h$counts, h$counts[length(h$counts)])
      lines(hx, hy/max(hy) * 0.95, lwd = 1, type = "s", 
            col = denscol)
      axis(2, at = pretty(hy)/max(hy) * 0.95, pretty(hy))
      title("Color Key\nand Histogram")
      par(cex = 0.5)
      mtext(side = 2, "Count", line = 2)
    }
    else title("Color Key")
    if (trace %in% c("both", "column")) {
      vline.vals <- scale01(vline, min.raw, max.raw)
      if (!is.null(vline)) {
        abline(v = vline.vals, col = linecol, lty = 2)
      }
    }
    if (trace %in% c("both", "row")) {
      hline.vals <- scale01(hline, min.raw, max.raw)
      if (!is.null(hline)) {
        abline(v = hline.vals, col = linecol, lty = 2)
      }
    }
  }
  else plot.new()
  retval$colorTable <- data.frame(low = retval$breaks[-length(retval$breaks)], 
                                  high = retval$breaks[-1], color = retval$col)
  if (!is.null(extrafun)) 
    extrafun()
  invisible(retval)
}
```

### 7.4 Decorating heatmaps
#### 7.4.1 Dendrograms

The other thing that determines the impression of heatmaps is the order
of the features and the samples. By default, R tries to make the picture
as clear as possible. This is achieved by re-ordering the features and
samples using dendrograms. Let's take a look at our heatmap without the
dendrogram and ordering:

```{r heatmap2-log-noScale-noDend}
heatmap.2a(metabo_log,
          dendrogram = "none", #no dendrogram plotted
          Rowv = F, #no sorting for rows
          Colv = F, #no sorting for columns
       col = hcl.colors(255,"YlGnBu", rev = TRUE), #single colour palette - 255 colours
        symbreaks = F, #colours are not centred around 0
        scale = "none", # no scaling!
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5), # remove the white space around
        trace = "none", #don't plot a trace
        density.info = "none") #don't overlay a histogram in the colour bar
```

It's hard to believe that these are the same values, right?

The default dendrogram is based on Euclidean distances, which are
hierarchically clustered using complete linkage. You can supply your own
dendrograms, based on your choice of distance:

```{r heatmap2-log-noScale-customDend}
heatmap.2a(metabo_log,
          dendrogram = "both", #both dendrograms plotted
          Rowv = as.dendrogram(hclust(dist(metabo_log,"minkowski"),
                        "ward.D2")), #custom
          Colv = as.dendrogram(hclust(dist(t(metabo_log)),
                        "ward.D2")), #custom dendrogram
       col = hcl.colors(255,"YlGnBu", rev = TRUE), #single colour palette - 255 colours
        symbreaks = F, #colours are not centred around 0
        scale = "none", # no scaling!
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5), # remove the white space around
        trace = "none", #don't plot a trace
        density.info = "none") #don't overlay a histogram in the colour bar
```

The choice of distance is up to you and depends on your data and what
you consider 'different'. Most fields of research have (more or less
justified) standards. \*Try out some of the options you can find under
`?dist` and `?hclust` to see effects.

#### 7.4.2 Ordering features and samples

We've already seen the power of sorting. Next to ordering you data based
on clustering (you might call this 'unsupervised'), you can have
pre-defined groups (especially for samples). You can also visualize
those:

```{r heatmap2-log-noScale-sampleOrd}
heatmap.2a(metabo_log[,order(metaD$Status,metaD$Gender,metaD$BMI)],# custom order using metadata: by study group, gender, and BMI
          dendrogram = "row", # row dendrogram plotted
         Colv = NULL, 
       col = hcl.colors(255,"YlGnBu", rev = TRUE), #single colour palette - 255 colours
        symbreaks = F, #colours are not centred around 0
        scale = "none", # no scaling!
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5), # remove the white space around
        trace = "none", #don't plot a trace
        density.info = "none") #don't overlay a histogram in the colour bar
```

Some structure is still visible. So, this looks like some of the
metabolites behave a bit differently in the samples on the left than on
the right. But the most obvious differences between samples we seemed to
see before may be due to something else. We can indicate the separation
of groups with vertical lines:

```{r heatmap2-log-noScale-sampleOrdL}
heatmap.2a(metabo_log[,order(metaD$Status,metaD$Gender,metaD$BMI)],# custom order using metadata: by study group, gender, and BMI
          dendrogram = "row", # row dendrogram plotted
          Colv = NULL, # no sorting (except above)
          colsep = cumsum(c(t(table(metaD$Status,metaD$Gender)))), #counting the grouping from above!
          col = hcl.colors(255,"YlGnBu", rev = TRUE), #single colour palette - 255 colours
          symbreaks = F, #colours are not centred around 0
          scale = "none", # no scaling!
          labRow = "", # no rownames
          labCol = "", # no colnames
          margins = c(0.5,0.5), # remove the white space around
          trace = "none", #don't plot a trace
          density.info = "none") #don't overlay a histogram in the colour bar
```

#### 7.4.3 Adding attributes

We can add some sample information to the heatmap using colours for a
similar effect:

```{r heatmap2-log-noScale-colSide}
heatmap.2a(metabo_log,
          ColSideColors = c("purple","green")[as.numeric(as.factor(metaD$Status))],
          dendrogram = "both", #both dendrograms plotted
       col = hcl.colors(255,"YlGnBu", rev = TRUE), #single colour palette - 255 colours
        symbreaks = F, #colours are not centred around 0
        scale = "none", # no scaling
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5), # remove the white space around
        trace = "none", #don't plot a trace
        density.info = "none") #don't overlay a histogram in the colour bar
```

Our improved `heatmap.2a` also takes multiple attributes for colour
codes.

```{r heatmap2-log-noScale-colSideM}
heatmap.2a(metabo_log,
          ColSideColors = rbind(c("pink","lightblue")[as.numeric(as.factor(metaD$Gender))],
                                brewer.pal(7,"Set3")[as.numeric(as.factor(metaD$Gastric.acid.medication))],
                                c("purple","green")[as.numeric(as.factor(metaD$Status))]),
          dendrogram = "both", #both dendrograms plotted
       col = hcl.colors(255,"YlGnBu", rev = TRUE), #single colour palette - 255 colours
        symbreaks = F, #colours are not centred around 0
        scale = "none", # no scaling
        labRow = "", # no rownames
        labCol = "", # no colnames
        margins = c(0.5,0.5), # remove the white space around
        trace = "none", #don't plot a trace
        density.info = "none") #don't overlay a histogram in the colour bar
```

There are more options to tweak. Take a look at `?heatmap.2` to see
them.
