---
title: "09 - Visualizing networks with igraph"
author: "Anna Heintz-Buschart"
date: '2022-09-10'
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: 4
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this notebook, we take a look at networks. R can handle networks,
both for analysis as well as for visualization. There are several
packages to help with this job. The methods in this tutorial are based
on the [igraph](https://igraph.org/r/) package. The igraph project
actually interfaces with several programming languages, so you may also
find it in other contexts.

### 9.1 Setup

```{r libraries}
if(!require(igraph)){
  install.packages("igraph",repos = "http://cran.us.r-project.org")
  library(igraph)
}
```

### 9.2 Network representations

There are multiple ways of reading in or defining a network. The two
most common ways to represent a network are *edge lists* and *adjacency
matrices*.

#### 9.2.1 Network formats: edgelist

In the first representation, you have a table with at least two columns.
Every line represents an edge. The two columns take the two nodes that
are connected in this line. Directionality is straight-forward with
this: You can define one column to contain only source nodes (where the
arrow starts) and the other column to contain only sink or target nodes
(where the arrow ends). Un-directed networks can be (implicitly or
explicitly) defined by giving every edge twice, with the source and sink
nodes of one line being exchanged in the other line. Directionality and
weights can also be defined in additional columns. A common input file
format is [simple interaction
format](http://www.cbmc.it/fastcent/doc/SifFormat.htm).

Here is the example of a simple network.

```{r exampleRing1}
edgelist_circle1 <- matrix(c("A","B","B","C","C","D","D","E","E","F","F","A"),
                           byrow = T,
                           ncol=2,
                           dimnames = list(c(1:6),c("source","sink")))
edgelist_circle1
```

```{r exampleRing1_igraph}
testgraph_circle1 <- graph_from_edgelist(edgelist_circle1)
plot(testgraph_circle1, layout = layout_in_circle(testgraph_circle1))
```

#### 9.2.2 Network formats: adjacency matrix

The second representation is a square matrix with one row and one column
each for every node. The fields that are spanned by this matrix contain
information on whether an edge connects the node belonging to this row
and the node belonging to this column. Usually, `0` indicates no
connection and `1` indicates a connection (weights can be introduced,
too, if necessary). Directionality is dealt with by using the two
triangles of the matrix: the matrix is read such that the rows indicate
source nodes and the columns sink nodes for the edges.

```{r exampleRing2}
#make an empty matrix with as many rows and columns as there are nodes
adjMat_circle2 <- matrix(0,
                         nrow=6,
                         ncol=6,
                         dimnames = list(c("J","K","L","M","N","O"),
                                         c("J","K","L","M","N","O")))
#fill matrix by connecting each node to only the next
for(i in 1:(nrow(adjMat_circle2)-1)){
  adjMat_circle2[i,i+1] <- 1
}
#close the circle
adjMat_circle2[nrow(adjMat_circle2),1] <- 1
adjMat_circle2
```

```{r exampleRing2_igraph}
testgraph_circle2 <- graph_from_adjacency_matrix(adjMat_circle2)
plot(testgraph_circle2, layout = layout_in_circle(testgraph_circle2))
```

**Exercise**: In the code below, change the edgelist in a way that
breaks the circle and adds a new edge instead. Plot the result.

```{r example_change1}
edgelist_new <- matrix(c("A","B","B","C","C","D","D","E","E","F","F","A"), #adapt this
                           byrow = T,
                           ncol=2,
                           dimnames = list(c(1:6),c("source","sink")))
edgelist_new
```

```{r example_change1_igraph}
testgraph_new1 <- graph_from_edgelist(edgelist_new)
plot(testgraph_new1, layout = layout_in_circle(testgraph_new1))
```

**Exercise**: In the code below, change the adjacency matrix in a way
that breaks the circle and adds a new edge instead. Plot the result.

```{r example_change2}
adjMat_new <- adjMat_circle2
#replace at least two fields by adjusting the following lines (you need to un-comment them by removing the hashkey):
# adjMat_new[x1,y1] <- 0
# adjMat_new[x2,y2] <- 1
adjMat_new 
```

```{r example_change2_igraph}
testgraph_new2 <- graph_from_adjacency_matrix(adjMat_new )
plot(testgraph_new2, layout = layout_in_circle(testgraph_new2))
```

### 9.3 Network visualization

base R and `igraph` are not really made for visualizing large
graphs/networks. There's better software outthere for those tasks. But
what is nice is that you can integrate your visualization with the
functions to analyse graphs and R's statistical tools and all the
visualizations for the results from those functions. So let's take a
look at `igraph`'s options for plotting graphs.

We'll first create a simple graph from a built-in function. In real
life, you'd be using an `igraph` object that you created from an
adjacency matrix or an edge list, depending on your data.

```{r atlasgraph}
g <- graph_from_atlas(501)
plot(g)
```

As you can see, this graph's nodes don't have names. Let's give them
some.

```{r names}
V(g)$name <- c("Maria","Johannes","Jan","Johanna","Cornelis","Hendrik","Anna")
plot(g)
```

#### 9.3.1 Layouts

Layouts are what determine where the nodes sit on the canvas. By
default, `igraph` guesses a good layout algorithm for your graph's size.
For a small graph, this is usually a Fruchterman-Reingold force-directed
layout, which will put more connected nodes closer to each other. It can
also be chosen explicitly:

```{r gfr}
plot(g,layout = layout_with_fr(g))
```

One important point to notice is that the layout does not always look
the same. If you run the above chunk a couple of times, you will end up
with slightly different positions of your nodes. This can be a problem,
if you want to plot the same graph twice, but highlight different
aspects. In this case, you can save a layout.

```{r saveLayout}
frg <- layout_with_fr(g)
```

Now, you can re-use this layout - run the following chunk a couple of
times, you will see that nothing changes:

```{r plotsavedLayout}
plot(g, layout=frg)
```

You've already met the circle layout above:

```{r gcircle}
plot(g,layout = layout_in_circle(g))
```

More layouts include stars:

```{r gstar}
plot(g,layout = layout_as_star(g))
```

... as trees (this one is nice for tree-like structures, e.g.
pedigrees):

```{r gtree}
plot(g,layout = layout_as_tree(g))
```

... other algorithms that try to bring more connected nodes together
while avoiding crowding, e.g.,

... using the Kamada-Kawai layout:

```{r gkk}
plot(g,layout = layout_with_kk(g))
```

... using the GEM layout:

```{r ggem}
plot(g,layout = layout_with_gem(g))
```

... or using multi-dimensional scaling of the distances between the
nodes in the graph (problematic with equi-distant nodes like we have
them here):

```{r gmds}
plot(g,layout = layout_with_mds(g))
```

... or just plain random:

```{r grand}
plot(g,layout = layout_randomly(g))
```

You can even set all the coordinates manually, by giving one coordinate
per node in a $n\times2$ matric (n being the number of nodes).

```{r gmanl}
plot(g,layout = matrix(sample(1:7,14,replace=T),nrow=7,ncol=2))
```

#### 9.3.2 Node labels

Let's keep our FR-layout from above and manipulate its look. *Exercise:*
play around with the visualization arguments shown below to understand
how they work. Let's start with the labels. You can choose not to plot
them:

```{r noLabels}
plot(g, layout=frg,
     vertex.label = NA) #no labels
```

In a very un-R-ish way, the default labels have serifs. You can change
to a more usual sans serif font like so:

```{r nsLabels}
plot(g, layout=frg,
     vertex.label.family = "sans") # sans serif labels
```

And change the color:

```{r colLabels}
plot(g, layout=frg,
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black") #color
```

... and the size:

```{r vlabSize}
plot(g, layout=frg,
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.7) # size
```

... also for the individual nodes:

```{r vlabSize2}
plot(g, layout=frg,
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.2*1:7) # size
```

With this, you could also take values from an analysis, e.g. the degree
(i.e. number of connections) of the nodes:

```{r vlabSize3}
plot(g, layout=frg,
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.2*degree(g)) # size based on degree
```

The labels don't need to sit in the node:

```{r vlabPos}
plot(g, layout=frg,
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.6, # size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

#### 9.3.3 Node shapes and sizes

You have a choice of different shapes, e.g. circles, squares,
rectangles, spheres.

```{r vShape}
plot(g, layout=frg,
     vertex.shape="square",
     vertex.size = 40,
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", # label color
     vertex.label.cex = 0.6) # label size 
```

As always, they can be set by node:

```{r vShape2}
plot(g, layout=frg,
     vertex.shape=c("square","circle")[1+as.numeric(grepl("nn",V(g)$name))],
     vertex.size = 40,
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", # label color
     vertex.label.cex = 0.6) # label size 
```

Colours of both the node and its frame can be chosen freely, too:

```{r vShape3}
plot(g, layout=frg,
     vertex.shape=c("square","circle")[1+as.numeric(grepl("nn",V(g)$name))],
     vertex.color = c("darkgreen","orange")[1+as.numeric(grepl("^J",V(g)$name))],
     vertex.frame.color = c("black","grey"), # vectors are not recycled, any node without a value gets an NA
     vertex.size = 40,
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", # label color
     vertex.label.cex = 0.6) # label size 
```

A special case of nodes are pies:

```{r vPies}
plot(g, layout=frg,
     vertex.shape= "pie",
     vertex.size = 30,
     vertex.pie = list(c(1,1,0,2),
                       c(2,1,0,1),
                       c(3,1,0,0),
                       c(4,0,1,0),
                       c(2,2,0,3),
                       c(1,1,0,1),
                       c(4,4,0,1),
                       c(5,0,0,1)),
     vertex.pie.color = list(rainbow(4)),
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", # label color
     vertex.label.cex = 0.6, # label size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

Which are especially effective with the size vector:

```{r vPies2}
pieList <- list(c(1,1,0,2),
                c(2,1,0,1),
                c(3,1,0,0),
                c(4,0,1,0),
                c(2,2,0,3),
                c(1,1,0,1),
                c(4,4,0,1),
                c(5,0,0,1))
plot(g, layout=frg,
     vertex.shape= "pie",
     vertex.size = sapply(pieList,sum)*6,
     vertex.pie = pieList,
     vertex.pie.color = list(rainbow(4)),
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", # label color
     vertex.label.cex = 0.6, # label size 
     vertex.label.dist = rep(3,7), # label distance to node centre
     vertex.label.degree = rep(0,7)) # label position 0=right, pi=left, pi/2=below etc
```

#### 9.3.4 Edge appearance

Of course, the edges can also be controlled individually.

```{r ecol}
plot(g, layout=frg,
     edge.color = "black",
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.6, # size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

... including the line type:

```{r elty}
plot(g, layout=frg,
     edge.lty = 3,
     edge.color = "black",
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.6, # size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

... and the thickness:

```{r ewidth}
plot(g, layout=frg,
     edge.width = 3,
     edge.lty = 2,
     edge.color = "black",
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.6, # size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

... which can, of course be adapted per edge:

```{r elty2}
plot(g, layout=frg,
     edge.width = sqrt(edge_betweenness(g)),
     edge.lty = 2,
     edge.color = "black",
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.6, # size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

Edges can be curved:

```{r ecurve}
plot(g, layout=frg,
     edge.curved = T,
     edge.width = sqrt(edge_betweenness(g)),
     edge.lty = 1,
     edge.color = "black",
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.6, # size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

... and more or less curved:

```{r ecurve1}
plot(g, layout=frg,
     edge.curved = 0.1,
     edge.width = sqrt(edge_betweenness(g)),
     edge.lty = 1,
     edge.color = "black",
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.6, # size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

```{r ecurve2}
plot(g, layout=frg,
     edge.curved = -2,
     edge.width = sqrt(edge_betweenness(g)),
     edge.lty = 1,
     edge.color = "black",
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.6, # size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

Of course, edges may have labels:

```{r eLab}
plot(g, layout=frg,
     edge.label = E(g),
     edge.label.family = "sans",
     edge.label.cex = 0.6,
     edge.label.color = "magenta",
     edge.label.font = 3,
     edge.curved = 0.1,
     edge.width = sqrt(edge_betweenness(g)),
     edge.lty = 1,
     edge.color = "black",
     vertex.label.family = "sans", # sans serif labels
     vertex.label.color = "black", #color
     vertex.label.cex = 0.6, # size 
     vertex.label.dist = 3, # label distance to node centre
     vertex.label.degree = 0) # label position 0=right, pi=left, pi/2=below etc
```

You can also add a title, box and other things. See `?igraph.plotting`
for further tips.
